---
title: "Documentation on blockdss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documentation on blockdss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blockdss)
```

## Introduction

dss.c.64.exe is a software tool running geostatistical algorithms. Among the algorithms available, dss.c.64.exe runs block sequential simulation using direct sequential simulation algorithm (Soares, 2000^[https://doi.org/10.1023/A:1012246006212]). The blockdss package is a wrapper for running dss.c.64.exe block direct sequential simulation as an R package, as it provides functions for interfacing with dss.c.64.exe.

This tutorial shows how to run block sequential simulation with R for disease mapping, modelling the spatial distribution of covid-19, as in Azevedo et al. 2020^[https://doi.org/10.1186/s12942-020-00221-5]. For that purpose the tutorial shows step-by-step guide using covid-19 data to simulate covid-19 risk maps, median risk map and risk uncertainty map. 

A wrapper function for running a visualization tool developed by Aimee Taylor and colleagues^[https://doi.org/10.48550/arXiv.2005.11993] is also included in blockdss package to merge covid-19 median risks and associated spatial uncertainty in a single-map.

## Software: dss.c.64.exe

In default (working) directory create a folder called `input` and put the dss.c.64.exe^[available at github] inside.

## Datasets

To perform the block simulation algorithm a minimum number of 2 datasets are required: 

- a data frame with spatial data point locations;  
- a grid with id region values at all simulation locations (*grid nodes*).

Datasets are available for this example in the blockdss package:

```{r}
data(ptdata)
data(ptgrid)
```

Write `?ptdata` and `?ptgrid` in console for details on datasets.

### Convert `ptgrid`

A transformation to `SpatialPixelsDataFrame`:
```{r ptgrid transform, warning = F, message = F}
library(sp)
coordinates(ptgrid) <- ~x+y
proj4string(ptgrid) <- CRS("+init=epsg:3763")
class(ptgrid)
gridded(ptgrid) <- T
class(ptgrid)
```


### `irates()` to compute municipality rates and variance-error terms

Syntax:
```{r args irates, echo = F}
str(args(irates))
```

Use `irates()` to compute rates, variance-error terms by region.
The arguments of the function are:

- `dfobj`, string, dataframe name with disease data
- `oid`, character, field name for region id
- `xx`, character, fieldname for x-coordinates 
- `yy`, character, fieldname for y-coordinates 
- `zz`, character, fieldname for z-coordinates 
- `cases`, character, fieldname for number of cases
- `pop`, character, fieldname for population size
- `casesNA`, numeric, an integer used to replace rows with cases = NA,
- `day`, character, string indicating date (format "yyyymmdd") of disease data
- `perhab`, numeric, an integer indicating how much the rate is multiplied for, to express the disease rate (e.g 100000)


```{r fun rates-var}
rates = irates(dfobj = ptdata, oid = "oid_", xx = "x", yy = "y", zz = "t", 
               cases = "ncases", pop = "pop19", casesNA = 1, day = "20210115")
```

The function `irates()` returns the following list of objects:
```{r rates-var-return}
str(rates)
```

It also writes a text file (.out) with rates and store it in `input` folder. 

### `blockfile()` to create a file with blocks (municipality areas) 

Syntax:
```{r args blockfile, echo = F}
str(args(blockfile))
```

Use `blockfile()` to transform grid file in block format. The function requires some libraries to be loaded. If not installed they will be first installed.

You should provide a georeferenced grid file with id region values at simulation locations.
The arguments of the function are:

- `rateobj`, string, name of list, output of function `irates()`, 
- `gridimage`, character,  name of grid file, 
- `na.value`, numeric, integer with grid value for "No data"  

```{r fun block-data, warning = F, message = F}
block = blockfile(rates, ptgrid)
```

The grid file values should refer to the region id's at simulation locations (nodes). All regions in covid data should be represented by 1 or more node. 

The function writes a text file (.out) with blockdata and store it in `input` folder. `blockfile()` also returns the following list of objects:

```{r block-data-return}
str(block, max.level = 2)
```

### `maskfile()` to create mask for block file.

Syntax:
```{r args maskfile, echo = F}
str(args(maskfile))
```

The function `maskfile()` creates a mask for the block file. The only argument of the function is the name of list, output of function `blockfile()`.

```{r fun mask-file}
mask = maskfile(block)
```

The function returns the following a list of objects and generates a text file (.out) with values {-1,0} where -1 are assigned to nodata locations and 0 are assigned to nodes with values (id region). The file is stored in `input` folder. 

```{r mask-file-return}
str(mask)
```

### `varexp()` computes experimental variogram

Syntax:
```{r args varexp, echo = F}
str(args(varexp))
```

Use `varexp()` to calculate experimental variogram from COVID-19 rates.
Only implemented in omnidirectional case.
The arguments are:

- `dfobj`, string, name of list, output of function `irates()`
- `lag`, numeric, the lag distance used for variogram estimates 
- `nlags`, numeric, the number of lags to calculate variogram.

```{r fun varexp}
vexp = varexp(rates, lag = 7000, nlags = 15)
```

The function returns a list with the weighted variance (by population size) and variogram estimates at nlags.

```{r varexp-str}
str(vexp)
```

```{r varexp-plot, out.width = "50%", fig.align = "center"}
plot(vexp[["semivar"]][1:2], ylab = expression(paste(gamma, "(h)")), xlab = "h (in m)") 
```

### `varmodel()` to fit a variogram model

Syntax:
```{r args varmodel, echo = F}
str(args(varmodel))
```

Funtion `varmodel()` fits (manually) a theoretical variogram. You should provide the experimental variogram data to evaluate fit by visual inspection, the variogram model type and the variogram parameters.
The arguments of `varmodel()` are:

- `varexp`, string, name of object, output of function varexp() 
- `mod`, character, the variogram model type (available are: "Sph" or "Exp")
- `nug`, numeric, nugget-effect value of the variogram
- `ran`, numeric, range value of the variogram
- `sill`, numeric, sill (or partial sill) value of the variogram

```{r fun varmodel}
vmod = varmodel(vexp, mod = "sph", nug = 0, ran = 35000, sill = vexp[["weightsvar"]])
```

`varmodel()` returns the following list of objects:

```{r varmodel-str}
str(vmod)
```

```{r varmodel-plot, out.width = "50%", fig.align = "center"}
# plot experimental variogram
plot(vexp[["semivar"]][1:2], ylab = expression(paste(gamma, "(h)")), xlab = "h (in m)") 
# add sill
abline(h = vexp[["weightsvar"]], col ="red", lty = 2)
# add theoretical model
lines(vmod[["fittedval"]]) 
```

### `ssdpars()` to create parameters file for dss.c.64.exe 

Syntax:
```{r args ssdpars , echo = F}
str(args(ssdpars))
```

Function `ssdpars()` generates a parameters file (.par) and the simulated maps in .out format. Function `ssdpars()` invokes dss.c.64.exe specified by the parameters file.

Arguments include names of lists and parameter values required for simulation processes:

- `blockobj`, string, name of list, output of function blockfile()
- `maskobj`, string, name of list, output of function maskfile()
- `dfobj`, string, name of list, output of function irates()
- `varmobj`, string, name of list, output of function varmodel()
- `simulations`, numeric, number of simulations
- `nrbias`, numeric, nr simulations for bias correction
- `biascor`, num vector, flag for (mean, variance) correction (yes = 1, no = 0)
- `ndMin`, numeric, min number of neighbour observations used in  kriging
- `ndMax`, numeric, max number of neighbour observations used in  kriging
- `nodMax`, numeric, max number of previously simulated nodes used in  kriging
- `radius1`, numeric, search radii in the major horizontal axe
- `radius2`, numeric, search radii in the axe orthogonal (horizontal) to radius1
- `radius3`, numeric, search radii in the vertical axe
- `ktype`, numeric, the kriging type to be used (available are: 0 = simple, 1 = ordinary)  

Note that this process may take a while, depending mostly on the number of simulation nodes and number of simulations specified.

```{r fun ssdir, eval = F}
ssdpars(blockobj = block, maskobj = mask, dfobj = rates, varmobj = vmod, 
        simulations = 5, radius1 = 35000, radius2 = 35000)
```

Both parameters file (.par) and simulations files (.out) are stored in `input` folder.

```{r fun outraster}
maps = outraster(block)
```

```{r str outraster obj}
str(maps, max.level = 3)
```

```{r spplot sims}
spplot(maps[["simulations"]])
```

```{r spplot etype, out.width = "50%", fig.align = "center"}
spplot(maps[["etype"]])
```

```{r spplot uncert, out.width = "50%", fig.align = "center"}
spplot(maps[["uncertainty"]])
```

## Acknowledgements

Manuel Ribeiro acknowledges the financial support of the CERENA (project FCT-UIDB/04028/2020) and Fundação para a Ciencia e Tecnologia (research contract IF2018-CP1384). Manuel Ribeiro gratefully acknowledge CERENA-IST/UL researchers Leonardo Azevedo, Maria João Pereira and Amilcar Soares for the code in Matlab and Fortran.

